Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-04-07 analysis/paper/paper.Rmd
removed reference to zenodo DOI... for now
minor edits from NP

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the fall DSR and spring NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. Our intent was to survey during the emigration of the 2018 Chinook Salmon brood year that occurred fall 2019 and spring 2020, however logistical constraints during the onset of the Covid-19 pandemic delayed the spring 2020 survey until 2021. Over this period, an additional fall survey occurred during the 2020 DSR emigration. Fall surveys were constrained to two weeks to minimize Northern Pikeminnow immigration and emigration.

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes occupying Deadwater Slough. Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks to be analyzed in a controlled environment. For each stomach sample, total wet weight (grams) was recorded for all stomach contents, including all fish and non-fish items (e.g., macroinvertebrates, organic matter). Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones [@Hansel1988; @Frost2000] or were categorized as unknown. Approximately 5% of Northern Pikeminnow were euthanized for dissection (n = 75) after gastric lavage to validate the efficacy of the methodology.

@@

All data and code for the analyses presented here can be found in a GitHub repository at ***zenodo DOI***.

@@

The lengths of Northern Pikeminnow ranged from `r min(cmr_df$length, na.rm = T)` to `r max(cmr_df$length, na.rm = T)` mm TL with an average of `r round(mean(cmr_df$length, na.rm = T))` mm (Figure \@ref(fig:length-freq-p)). The PSD for Northern Pikeminnow in Deadwater Slough across all three surveys was `r prettyNum(psd, digits = 1)`%.

@@

Given yearly fish consumption estimates during the fall DSR and spring NRR periods, we produced a sensitivity analysis to calculate the consumption of juvenile Chinook Salmon where 30%, 50%, and 65% of total fish prey consumed are juvenile Chinook Salmon (Figure \@ref(fig:juv-con-p)). At the median value of 50%, we estimated that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$juv_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$juv_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) juvenile Chinook Salmon will be consumed. Using the median Granite-to-Granite SAR for Chinook Salmon in the Upper Salmon River, we estimated the "adult equivalents" of juveniles consumed to be `r prettyNum(round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$adult_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$adult_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) adults (Figure \@ref(fig:adult-imp-p)).

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupies Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration even after accounting for the potential of delayed mixing of marked fish. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledged that their methods were unsuitable for sampling offshore, noting that water velocity, depth, and irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, the maximum depth at Deadwater Slough was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimated densities of Northern Pikeminnow suggest that slow-water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available throughout the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the basin; additional information on contemporary piscine predator abundances may be needed [@Widener2021].

Several assumptions in our estimators may have influenced the magnitude of our abundance results. First, all models assumed a closed population; however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Emigration of marked individuals could reduce the marking fraction in the population leading to an upward bias in abundance estimates, but emigration rates would need to be substantial. The multiple census estimators, which we report, are more robust to this assumption because the marking fraction is estimated daily. We additionally have little reason to believe that emigration occurs during the two-week survey window owing to favorable habitat for Northern Pikeminnow in Deadwater Slough relative to adjacent reaches. Given the size of the sampling area and the short duration of our surveys, the closed population assumption is likely met. This assumption can likewise affect results if mortality occurs for some marked fish released back to the population. During field processing, no mortalities or injury post release was observed and fish injured prior to release were retained. Nevertheless, the potential for mortality amongst marked and released fish cannot be discounted.

@@

In addition to the large population abundance, the observed (`r round(psd)`%) PSD in this study was greater than observations of 41% below Bonneville Dam and 18% in Bonneville Reservoir [@Winther2020]. This suggests that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. Notably, the Idaho state catch-and-release record Northern Pikeminnow, measuring `r max(gastric_df$length, na.rm = T)` mm TL, was caught in Deadwater Slough during the fall 2020 survey.

@@

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids in the Columbia River [@Winther2020]. A local Northern Pikeminnow bounty program could provide monetary incentive for sportfisherman, in addition to boosting the local economy from lost revenues. Bounty programs could be conducted year-round or seasonally to coincide with peak juvenile outmigration(s). Each approach would require continual/annual effort and harvest to suppress the population size and as such, appropriate funding would be needed to support the program. Lastly, a more passive approach could be to provide cover or structure within Deadwater Slough that provides refugia for juvenile salmonids during their emigration. Cover should be appropriately sized to provide concealment to juvenile fishes while reducing access by larger fishes like Northern Pikeminnow. Adding cover could be a cost-effective approach; however, its potential effectiveness is unclear.

@@

```{r juv-con-p, fig.cap = "Estimated number of juvenile Chinook Salmon consumed by Northern Pikeminnow from a sensitivity analysis of variable diet scenarios. The proportion of Chinook Salmon in the total fish prey consumed by Northern Pikminnow were evaluated for 30%, 50%, and 65% scenarios, similar to those reported by Zimmerman and Ward 1999 "}

>>>>

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the fall DSR and spring NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. Our intent was to survey during the emigration of the 2018 Chinook Salmon brood year that occurred fall 2019 and spring 2020, however logistical constraints during the onset of the COVID-19 pandemic delayed the spring 2020 survey until 2021. Over this period, an additional fall survey occurred during the 2020 DSR emigration. Fall surveys were constrained to two weeks to minimize Northern Pikeminnow immigration and emigration.

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes occupying Deadwater Slough. Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks to be analyzed in a controlled environment. For each stomach sample, total wet weight (grams) was recorded, including all fish and non-fish items (e.g., macroinvertebrates, organic matter). Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones [@Hansel1988; @Frost2000] or were categorized as unknown. Approximately 5% of Northern Pikeminnow were euthanized for dissection (n = 75) after gastric lavage to validate the efficacy of the methodology. Additionally, contents of livewells were examined during fish processing to ensure food items were not regurgitated.

@@

@@

Lengths of Northern Pikeminnow ranged from `r min(cmr_df$length, na.rm = T)` to `r max(cmr_df$length, na.rm = T)` mm TL with an average of `r round(mean(cmr_df$length, na.rm = T))` mm (Figure \@ref(fig:length-freq-p)). The PSD for Northern Pikeminnow in Deadwater Slough across all three surveys was `r prettyNum(psd, digits = 1)`%.

@@

Given yearly fish consumption estimates during the fall DSR and spring NRR periods, we produced a sensitivity analysis to calculate the consumption of juvenile Chinook Salmon where 30%, 50%, and 65% of total fish prey consumed are juvenile Chinook Salmon (Figure \@ref(fig:juv-con-p)). At the median value of 50%, we estimated that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$juv_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$juv_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) juvenile Chinook Salmon would be consumed. Using the median Granite-to-Granite SAR for Chinook Salmon in the Upper Salmon River, we estimated the "adult equivalents" of juveniles consumed to be `r prettyNum(round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$adult_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$adult_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) adults (Figure \@ref(fig:adult-imp-p)).

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupies Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration even after accounting for the potential of delayed mixing of marked fish. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledged that their methods were unsuitable for sampling offshore, noting that water velocity, depth, and irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, the maximum depth at Deadwater Slough was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimated densities of Northern Pikeminnow suggest that slow-water reaches outside of reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available throughout the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the basin; additional information on contemporary piscine predator abundances may be needed [@Widener2021].

Several assumptions in our estimators may have influenced the magnitude of our abundance results. First, all models assumed a closed population; however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Emigration of marked individuals could reduce the marking fraction in the population leading to an upward bias in abundance estimates, but emigration rates would need to be substantial. We also captured two fish with marks from previous years, indicating seasonal residency of some fish. The multiple census estimators, which we report, are more robust to this assumption because the marking fraction is estimated daily. We additionally have little reason to believe that emigration occurs during the two-week survey window owing to favorable habitat for Northern Pikeminnow in Deadwater Slough relative to adjacent reaches. Given the size of the sampling area and the short duration of our surveys, the closed population assumption is likely met. This assumption can likewise affect results if mortality occurs for some marked fish released back to the population. During field processing, no mortalities or injury post release was observed and fish injured prior to release were retained. Nevertheless, the potential for mortality amongst marked and released fish cannot be discounted.

@@

In addition to the large population abundance, the observed (`r round(psd)`%) PSD in this study was greater than observations of 41% below Bonneville Dam and 18% in Bonneville Reservoir [@Winther2020]. This suggests that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. Notably, the Idaho state catch-and-release record Northern Pikeminnow, measuring `r max(gastric_df$length, na.rm = T)` mm TL, was caught in Deadwater Slough during the fall 2020 survey alluding to the favorable conditions for Northern Pikeminnow.

@@

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids in the Columbia River [@Winther2020]. A local Northern Pikeminnow bounty program could provide monetary incentive for sportfisherman, in addition to boosting the local economy from lost revenues. Bounty programs could be conducted year-round or seasonally to coincide with peak juvenile outmigration(s). Each approach would require continual/annual effort and harvest to suppress the population size and as such, appropriate funding would be needed to support the program. Lastly, a more passive approach could be to provide cover or structure within Deadwater Slough that provides refugia for juvenile salmonids during their emigration. Cover would need to be appropriately sized to provide concealment to juvenile fishes while reducing access by larger fishes like Northern Pikeminnow. Adding cover could be a cost-effective approach; however, its potential effectiveness is unclear.

@@

```{r juv-con-p, fig.cap = "Estimated number of juvenile Chinook Salmon consumed by Northern Pikeminnow from a sensitivity analysis of variable diet scenarios. The proportion of Chinook Salmon in the total fish prey consumed by Northern Pikminnow were evaluated for 30%, 50%, and 65% scenarios, similar to those reported by Zimmerman and Ward 1999."}
